Fast gravitational wave parameter estimation with no-compromise

### 2023-01-22 src/tex/ms.tex
Update Heterodyne likelihood section

% Title
\title{Turbo fast no compromise gravitational wave parameter estimation}
% Title Options
% \title{Ripples in the Flow: \\ Fast, General Parameter Estimation with Normalizing Flows and Differentiable Waveforms}

@@

% Abstract with filler text
\begin{abstract}
We present a lightweight, flexible, and high-performance framework for
inferring the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible
waveforms, and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced
by normalizing flows, we achieve full Bayesian parameter estimation for real
events like GW150914 and GW170817 within a minute of sampling time.  \mi{should
we say a standard analysis takes hours?} Our framework does not require
pre-training or explicit reparameterizations and can be generalized to handle
higher dimensional problems. We present the details of our implementation and
discuss trade-offs and future developments in the context of other proposed
strategies for real-time parameter estimation. Our code for generating the
manuscript and running the analysis is publicly available on GitHub.
\end{abstract}

@@

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational-wave (GW) data analysis, and it underpins all of GW physics
and astrophysics \cite{Christensen:2022bxb,
2019PASA...36...10T}. The central goal of PE is to infer the parameters of a
particular GW source given the strain data detected by instruments like LIGO \cite{LIGOScientific:2014pky}, Virgo \cite{VIRGO:2014yos}
and KAGRA \cite{KAGRA:2020tym}.
In the standard compact binary coalescence (CBC) scenario, this
could mean inferring intrinsic parameters such the masses and spins of the
compact objects, as well as extrinsic parameters such as their sky localization and distance
from Earth. PE is also applied to test general relativity (GR) and constrain deviations away from its predictions in observed data
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary evolution
\cite{LIGOScientific:2021psn} and measuring the properties of nuclear matter.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including \textsc{LALInference}
\cite{Veitch:2014wba}, \textsc{PyCBC Inference} \cite{Biwer:2018osg}, and \textsc{Bilby}
\cite{Ashton:2018jfp}.
These packages have been tested by a number of groups and
are well regarded as the standard tools. However, while these tools have passed
many robustness tests, they are known to be computationally intensive. The exact
amount of time needed to analyze one event depends on factors like the duration and frequency of the signal, as well as features of the specific waveform model.
Typical runtimes for production-level analyses of shorter events, such as binary
black hole (BBH) mergers, can range from hours to weeks, whereas longer events such as binary neutron star (BNS) mergers can reach months on a standard computer cluster.
\mi{do we have any reference for these numbers? this will be attacked otherwise, since the comparisons are not strictly fair---e.g., it takes me 2 days to run on GW170817 with IMRD, not months}
This expense precludes iterating quickly on results, launching large scale measurement simulations, or obtaining results in low latency to inform astronomers for potential followup in real time.

Additionally, in
the coming decade, there are planned upgrades for existing facilities, as well as
plans for next-generation detectors such as the Einstein Telescope (ET)
\cite{Punturo:2010zz} and the Cosmic Explorer (CE)
\cite{LIGOScientific:2016wof}. These upgrades will increase the sensitivity of
the instruments and allow for the detection of more events with better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \cite{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools.

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups developing
tools to speed up the PE process. This includes methods that employ modern
tools such as deep learning-based method pre-trained on a large collection of
waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that simplify the
computational challenge by leveraging our knowledge about GW signals
\cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising avenues
to tackle standard GW analysis tasks in the coming decade, particularly about
events involving CBCs in GR, they rely on assumptions that may not be valid for
analysis involving additional physical effects such as lensing and beyond-GR
analysis, or may use approximations that do not compute the Bayesian likelihood
exactly.

@@

In this work, we present a lightweight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific models
\mi{what do you mean by ``models'' here? we rely on specific waveforms} or
assumptions about the problem to achieve its performance. This makes our method
extensible to problems beyond the standard CBC analysis, without sacrificing
accuracy for efficiency.
\mi{should we also say something of the advantage of having better, modern
sampling with diffable WFs? can we say we expect this to be more robust to
strange posterior geometries, as you might get by introducing tidal and non-GR
params?}

@@

The main objective in PE is to obtain a multidimensional posterior distribution
$p(\mathbf{\theta} \mid d)$ on parameters $\mathbf{\theta}$ given strain
data $d$.  Such probability density represents our best inference of
the source properties, an encodes all relevant information contained in the
observed data.
To compute this object, we use Bayes' theorem to write
\begin{align} \label{eq: bayes}
    p(\theta \mid d) = \frac{\mathcal{L}(d \mid \theta)\pi(\theta)}{p(d)}\, ,
\end{align}
where $\mathcal{L}(d \mid \theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins, or it could encode astrophysical information.
Assuming the noise is drawn from a Gaussian process, the
log-likelihood for GW data is given by
\begin{align}
    \log{\mathcal{L}(d \mid \theta)} = -\frac{1}{2} \left\langle d-h(\mathbf{\theta}) \mid d-h(\mathbf{\theta})\right\rangle,
\label{eq: loglikelihood}
\end{align}
where $d$ is the observed strain data, $h(\mathbf{\theta})$ is the signal predicted by a
waveform model with a specific set of source parameters $\theta$. The right hand side of
Eq.~\eqref{eq: loglikelihood} can be evaluated in either the time or frequency domains.
For stationary noise, it is computationally cheaper to compute the likelihood in the frequency
domain, and the noise-weighted inner product can be written 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left\langle a \mid b\right\rangle = 4 \Re \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)}\, \mathrm{d}f \, ,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD).
In practice, the integral becomes a discrete sum over a finite number of
samples determined by the sampling rate of the detector data and duration of
the observation.

@@

We now give a concise description of the implementation of this approach in our
code; for a more extensive derivation of heterodyned likelihood, we refer the
reader to the reference \cite{Zackay:2018qdy}. In the heterodyned likelihood
framework, the two terms involving $h$ obtained by expanding Eq.~\eqref{eq: loglikelihood} can be approximated as
\begin{subequations} \label{eq: heterodynedlikelihood}
\begin{equation}
    \langle d \mid h \rangle \approx \sum_b \left[ A_0(b)\, r^*_0(h,b) + A_1(b)\, r^*_1(h,b) \right] ,
\end{equation}
\begin{align}
    \langle h \mid h \rangle \approx \sum_b &\left[ B_0(b)\, |r_0(h,b)|^2 + \right. \nonumber \\
    &\left. 2 B_1(b)\, \Re\{r_0(h,b)\, r_1(h,b)\} \right] 
\end{align}
\end{subequations} 
where $b$ denotes the index of a \textit{sparse} set of bins over which the integrand
will be computed; $A_0(b)$, $A_1(b)$, $B_0(b)$, and $B_1(b)$ are the
heterodyning coefficients computed using the data and the reference waveform;
and, finally, $r_0(h,b)$ and $r_1(h,b)$ are the ratios between the target waveform and the
reference waveform at the center of the bin \mi{mention difference between $r_{0/1}$?}.
To evaluate Eq.~\eqref{eq:
heterodynedlikelihood}, we need to first choose a binning scheme,
then evaluate the coefficients given the data and the reference waveform, and last the
ratio between the target waveform and the reference waveform at the center of
each bin.
\mi{this paragraph is kind of floating around, maybe it should go at the end of
this subsection?}

We can obtain further insight by noting that the Fourier phase of a GW inspiral
can be expressed as a power series $\Psi(f) = \sum_i \alpha_i f^{\gamma_i}$,
where $\alpha_i$ are some coefficients depending on the waveform parameters and
$\gamma_i$ are powers motivated by post-Newtonian theory \cite{}. For example,
for the term $\gamma_i = -5/3$, $\alpha_i$ is related to the chirp mass. The
maximum dephasing one can accrue over a frequency interval
$[f_{\textrm{min}},f_{\textrm{max}}]$ is thus given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i \geq 0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in Eq.~\eqref{eq:
maxdephasing}, we can choose the binning scheme to divide the entire
frequency band of interest into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.,
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

For sufficiently fine bins, the ratio between the target waveform
and the reference within a bin can be approximated by linear interpolation,
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
\mi{I think this eqn should appear right after Eq.~\eqref{eq:
heterodynedlikelihood}, because it's necessary to understand what the terms in
that eqn are, and you don't need Eq.~\eqref{eq: maxdephasing} to understand this one}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin respectively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and inverting Eq.~\eqref{eq:definer}.

>>>>

% Title
\title{Fast  gravitational wave parameter estimation with no-compromise}
% Title Options
% \title{Ripples in the Flow: \\ Fast, General Parameter Estimation with Normalizing Flows and Differentiable Waveforms}

@@

% Abstract with filler text
\begin{abstract}
We present a lightweight, flexible, and high-performance framework for inferring
the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible waveforms,
and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced by
normalizing flows, we achieve full Bayesian parameter estimation for real events
like GW150914 and GW170817 within a minute of sampling time.  Our framework does
not require pre-training or explicit reparameterizations and can be generalized
to handle higher dimensional problems. We present the details of our
implementation and discuss trade-offs and future developments in the context of
other proposed strategies for real-time parameter estimation. Our code for
generating the manuscript and running the analysis is publicly available on
GitHub.
\end{abstract}

@@

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational-wave (GW) data analysis and it underpins all of GW
physics and astrophysics \cite{Christensen:2022bxb, 2019PASA...36...10T}. The
central goal of PE is to infer the parameters of a particular GW source given
the strain data detected by instruments like LIGO \cite{LIGOScientific:2014pky},
Virgo \cite{VIRGO:2014yos} and KAGRA \cite{KAGRA:2020tym}. In the standard
compact binary coalescence (CBC) scenario, this could mean inferring intrinsic
parameters such as the masses and spins of the compact objects, as well as
extrinsic parameters such as their sky localization and distance from Earth. PE
is also applied to test general relativity (GR) and constrain deviations away
from its predictions in observed data
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary
evolution \cite{LIGOScientific:2021psn} and measuring the properties of nuclear
matter.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \cite{Veitch:2014wba}, \textsc{PyCBC Inference}
\cite{Biwer:2018osg}, and \textsc{Bilby} \cite{Ashton:2018jfp}. These packages
have been tested by a number of groups and are well regarded as the standard
tools. However, while these tools have passed many robustness tests, they are
known to be computationally intensive. The exact amount of time needed to
analyze one event depends on factors like the duration and frequency of the
signal, as well as features of the specific waveform model. Typical runtimes for
production-level analyses can range from days to weeks. This expense precludes
iterating quickly on results, launching large-scale measurement simulations, or
obtaining results in low latency to inform astronomers for potential follow-up in
real time.

Additionally, in the coming decade, there are planned upgrades for existing
facilities, as well as plans for next-generation detectors such as the Einstein
Telescope (ET) \cite{Punturo:2010zz} and the Cosmic Explorer (CE)
\cite{LIGOScientific:2016wof}. These upgrades will increase the sensitivity of
the instruments and allow for the detection of more events with a better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \cite{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools.

% Effort going to new tools development.
To address this, there are efforts from multiple groups developing
tools to speed up the PE process. This includes methods that employ modern
tools such as deep learning-based methods pre-trained on a large collection of
waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that simplify the
computational challenge by leveraging our knowledge of GW signals
\cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising avenues
to tackle standard GW analysis tasks in the coming decade, particularly about
events involving CBCs in GR, they rely on assumptions that may not be valid for
analysis involving additional physical effects such as lensing and beyond-GR
analysis, or may use approximations that do not compute the Bayesian likelihood
exactly.

@@

In this work, we present a lightweight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific assumptions
about the problem to achieve its performance. This makes our method extensible
to problems beyond the standard CBC analysis, without sacrificing accuracy for
efficiency.

@@

The main objective of PE is to obtain a multidimensional posterior distribution
$p(\mathbf{\theta} \mid d)$ on parameters $\mathbf{\theta}$ given strain
data $d$.  Such probability density represents our best inference of
the source properties, an encodes all relevant information contained in the
observed data.
To compute this object, we use Bayes' theorem to write
\begin{align} \label{eq: bayes}
    p(\theta \mid d) = \frac{\mathcal{L}(d \mid \theta)\pi(\theta)}{p(d)}\, ,
\end{align}
where $\mathcal{L}(d \mid \theta)$ is the likelihood function, $\pi(\theta)$ is
the prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins, or
it could encode astrophysical information. Assuming the noise is drawn from a
Gaussian process, the log-likelihood for GW data is given by
\begin{align}
    \log{\mathcal{L}(d \mid \theta)} = -\frac{1}{2} \left\langle d-h(\mathbf{\theta}) \mid d-h(\mathbf{\theta})\right\rangle,
\label{eq: loglikelihood}
\end{align}
where $d$ is the observed strain data, $h(\mathbf{\theta})$ is the signal
predicted by a waveform model with a specific set of source parameters $\theta$.
The right hand side of Eq.~\eqref{eq: loglikelihood} can be evaluated in either
the time or frequency domains. For stationary noise, it is computationally
cheaper to compute the likelihood in the frequency domain and the
noise-weighted inner product can be written 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left\langle a \mid b\right\rangle = 4 \Re \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)}\, \mathrm{d}f \, ,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD).
In practice, the integral becomes a discrete sum over a finite number of
samples determined by the sampling rate of the detector data and duration of
the observation.

@@

We now give a concise description of the implementation of this approach in our
code; for a more extensive derivation of heterodyned likelihood, we refer the
reader to the reference \cite{Zackay:2018qdy}. In the heterodyned likelihood
framework, the two terms involving $h$ obtained by expanding Eq.~\eqref{eq:
loglikelihood} can be approximated as
\begin{subequations} \label{eq: heterodynedlikelihood}
\begin{equation}
    \langle d \mid h \rangle \approx \sum_b \left[ A_0(b)\, r^*_0(h,b) + A_1(b)\, r^*_1(h,b) \right] ,
\end{equation}
\begin{align}
    \langle h \mid h \rangle \approx \sum_b &\left[ B_0(b)\, |r_0(h,b)|^2 + \right. \nonumber \\
    &\left. 2 B_1(b)\, \Re\{r_0(h,b)\, r_1(h,b)\} \right] 
\end{align}
\end{subequations} 
where $b$ denotes the index of a \textit{sparse} set of bins over which the integrand
will be computed; $A_0(b)$, $A_1(b)$, $B_0(b)$, and $B_1(b)$ are the
heterodyning coefficients computed using the data and the reference waveform;
and, finally, $r_0(h,b)$ and $r_1(h,b)$ are the ratios between the target waveform and the
reference waveform at the center of the bin and its first derivative.
For sufficiently fine bins, the ratio between the target waveform
and the reference within a bin can be approximated by linear interpolation,
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin respectively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and inverting Eq.~\eqref{eq:definer}.
To evaluate Eq.~\eqref{eq: heterodynedlikelihood}, we need to first choose a
binning scheme, then evaluate the coefficients given the data and the reference
waveform, and at last the ratio between the target waveform and the reference
waveform at the center of each bin.

Considering the phasing of a waveform is denoted by a power series $\Psi(f) =
\sum_i \alpha_i f^{\gamma_i}$, where $\alpha_i$ are some coefficients depending
on the waveform parameters and $\gamma_i$ are powers motivated by post-Newtonian
theory. For example, for the term $\gamma_i = -5/3$, $\alpha_i$ is related to
the chirp mass. The maximum dephasing one can have within a frequency interval
$[f_{\textrm{min}},f_{\textrm{max}}]$ is given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i \geq 0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in Eq.~\eqref{eq:
maxdephasing}, we can choose the binning scheme to divide the entire frequency
band of interest into a set of bins such that the maximum dephasing within each
bin is smaller than a certain threshold $\epsilon$, i.e.,
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.
